 

### 2021-05-12 vertical-thermal-review.Rmd
still, 688 words over

**Light conditions, specifically changes in the proportion of incident light and photosynthetically active radiation (PAR), vary along the vertical gradient with leaf area index (LAI), canopy height, canopy structure, and across species and forest types [Fig. `r fig_NEON_vertical`; @koike_leaf_2001].** 
Generally, along a vertical gradient in forests globally, amount of light decreases from canopy tops to the forest floor. This trend is mediated by foliage LAI and density, where the gradient is more prounced in broad-leaf forests than needle-leaf forests [Fig. `r fig_NEON_vertical`; @lowmanForestCanopies1995; @aussenacInteractionsForestStands2000a; @vonarxSpatiotemporalEffectsForest2012a; @poorterMetaanalysisPlantResponses2019b; Chazdon & Fetcher, 1984; Smith et al., 2019; Tymen et al., 2017].Light reaching forest floors that is incident to the canopy top can range from approximately 1-2% in tropical forests  [@robertsStomatalBoundaryLayerConductances1990a; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019], 4-5% in mixed-wood boreal and northern temperate forests[@bartemucciOverstoryInfluencesLight2006] to 17-34% in needle-leaf forests [@baldocchiSeasonalVariationEnergy1997]. Small canopy gaps and wind-induced movement of canopy leaves and branches also contribute sunflecks to shaded forest regions that receive ~1/3 of their carbon during photosynthesis through these short light exposures [@waySunflecksTreesForests2012]. Additionally, in heterogeneous canopies [with high gap fractions and large variation in tree height], the distance from the outer canopy to the inner canopy is a better proxy for light environment than height (Parker 1995).

### Wind speeds 
**Wind speeds are also higher at the top of the canopy (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`), owing to... [@juckerCanopyStructureTopography2018].**
Across all **#** forested NEON sites, wind speeds increased with height, with daily maximum wind speeds at the top of the averaging up to **#** m s^-1^ higher at the top of the vertical profile than at the bottom. Consistent with these results from NEON sites, higher wind speeds at greater heights within a forest canopy have been observed in both closed canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregor_tree_nodate] and open savannas [@curtis_intracanopy_2019].

@@

The strength of this buffering is greater under dense-canopy than open-canopy forests [*Defrenne et al. 2021*]. Where dense forests maintain daytime cooler lower maximum  temperature, and nighttime warmer higher minimum in the understory, relatively more than pine and boreal forests, nearby clearings [@vonarxSpatiotemporalEffectsForest2012a, *Defrenne et al. 2021*; @davis_microclimatic_2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009] and neighboring tree crop agricultural plantations [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018]. On the other hand, open-forests maintain higher daytime maximum temperature in the understory closer to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a; @martin_boundary_1999; @ramboCanopyMicroclimateResponse2009] and  sometimes warmer nighttime understory temperatures than above canopy, due to long wave infrared radiation from the soil [@vonarxSpatiotemporalEffectsForest2012a].

Seasonally, buffering capacity increases with water availablity [@davis_microclimatic_2019] and is greater in wet than dry season, in tropical and temperate forests [@rey-sanchez_spatial_2016; @vonarxSpatiotemporalEffectsForest2012a; @fauset_differences_2018, Tymen et al. 2017; @mcgregor_tree_nodate]. However, under colder conditions, dense forest canopies maintain warmer higher minimum temperatures [$T_{min}$] in the understory relative to open-forests and nearby clearings [@zellweger_seasonal_2019; *Defrenne et al. 2021*; @nakamuraForestsTheirCanopies2017; @mcgregor_tree_nodate] where greater radiative heat loss in exposed areas results in lower $T_{min}$.

Tree height also plays a role in buffering where trees upto 20m with higher LAI have shown to lower mean and max temperature and VPD in the understory. However, beyond a certain tree height, temperature and VPD offset can plateau [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018].

### Humidity and evaporative demand
**Humidity also varies across the forest vertical profile, and is generally higher in the understory (Fig. `r fig_NEON_vertical`).**[@mcgregor_tree_nodate]
Along with temperature, dense-canopy forests maintain higher daily maximum relative humidity [RH] below-canopy than open forests and nearby open areas. This trend is greater in wetter conditions, in summer and autumn months than winter [@vonarxSpatiotemporalEffectsForest2012a]. RH, in combination with temperature, determines vapor pressure deficit (VPD), an important metric expressing the driving force of water loss from a leaf.  Tropical forests during the wet season tend to have higher VPD, therefore greater evaporative demand, in the upper canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].

@@

With altitude, the strength of buffering decreases, moving to open-forests  [[@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a; @rajsnerovaMorphologicalBiochemicalPhysiological2015; @liangForestTypeHeight2019], increases in wetter than drier regions [**REF**] and with proximity to water, for eg. in riparian forests. [@davisMicroclimaticBufferingForests2019; @macekMaximumAirTemperature2019]

@@

**Many of the vertical gradient biophysical and trait variables reviewed above affect leaf temperature, $T_{leaf}$, which in turn has a strong influence on leaf metabolism (see following section).**

Leaves adapt to their biophysical environment to maximize carbon assimilation, by maintaining optimum leaf temperature in a given circumstance [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020]. As a result, leaves can exhibit 'limited homeothermy' where they are cooler than air at high temperature and warmer than air at low temperature or 'megathermy' where they are drastically warmer than $T_{air}$. In some cases, 'poikilothermy', where $T_{leaf}$ can be equalled to $T_{air}$ [@cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020].

$T_{leaf}$ is determined by the energy balance of a leaf, influenced by leaf traits, and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016] (Fig. `r fig_leaf_T`). The difference between $T_{leaf}$ and $T_{air}$ is determined as a function of energy input (through radiation or heat) versus heat lost (such as transpiration) to the environment [Fig. 2; @michaletzPlantThermoregulationEnergetics2015].

@@

Leaf thermoregulation is a tradeoff between water use strategy and optimum leaf temperature [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994]. With sufficient stomatal conductance ($g_s$) leaves can regulate their temperature, by dissapating heat, even in drastic biophysical environment [**REF**] but when leaf water demand exceeds water transport, driven by irradiance, $T_{air}$, VPD or drought, conservation of water ($gs$ limitation) occurs at the cost of increasing $T_{leaf}$.

@@

Greater upper canopy wind speed (Fig. `r fig_NEON_vertical`) decreases $T_{leaf}$ - $T_{air}$ by increasing boundary layer conductance ($g_a$) for exposed leaves [@daudetWindSpeedLeaf1999a]. Therefore, under hot conditions, leaves exposed to higher wind speeds would be most effective at cooling with high $g_s$ [Drake et al. 2018]. At low wind conditions, small $g_a$ to heat transfer can increase $T_{leaf}$ - $T_{air}$ [@martin_boundary_1999; @leighInfluenceLeafSize2017] where boundary layer resistance increases with thickness of air layer at the leaf surface. This tends to be greater in buffered leaves [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990a; @stokesBoundaryLayerConductance2006]

**(paragraph on hydraulics as a driver of leaf T:e.g., water on leaves - evaporative cooling - and leaf water content)**
While stomatal conductance is inherently greater in exposed leaves than buffered shaded leaves, factors driving $g_s$ limitation--irradiance, VPD and $T_{air}$--are also greater with exposure. As a result, exposed leaves can experience steeper $g_s$ limitation than buffered leaves, driving an increase in $T_{leaf}$, further influenced by irradiance and $T_{air}$ (@leighInfluenceLeafSize2017; @fauset_differences_2018). On the other hand, due to inherently lower $g_s$ in shade leaves, they can remain cooler under buffered conditions and substantially warmer under exposed conditions.

@@

### In-situ observations
**Vertical profiles in microclimate (Fig. `r fig_NEON_vertical`), leaf traits (see following section), and *evapotranspiration* (*where is this discussed? and do we say anything about evaporation?*) together shape in-situ leaf temperatures according to the principles outlined above (Fig. `r fig_leaf_T`).** 
Vertical distribution of leaf temperature is dependent on canopy buffering capacity and the biophysical gradient.

Particularly, open-forests with buffering capacity can have greater lower canopy $T_{leaf}$ than upper canopy, reflecting microclimate conditions of vertical air mixing, heating from the ground, light permeating through canopy layers and still air in the lower canopy [@mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987;@rey-sanchez_spatial_2016]. In other conditions, increase in inner canopy relative humidity can increase $T_{leaf}$  relatively more than above canopy $T_{leaf}$ [@dietzVerticalPatternsDuration2007]. Similarly, open trees growing close to the ground can experience greater heat stress and $T_{air}$ in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019]. Under other conditions $T_{leaf}$ tracks $T_{air}$ and therefore differs little between understory and canopy (Fig. `r fig_NEON_vertical`).

**However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.** In closed-canopy forests with increased buffering capacity, large $T_{leaf} - T_{air}$ increases with height where higher $T_{leaf}$ are found in upper than lower canopies, due to midday stomatal closure [Slot et al. 2019]. This is seen in tropical forests [@fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a], semi-deciduous tropical forest during the wet season [@rey-sanchez_spatial_2016], and as well in a temperate forest [@mauTemperateTropicalForest2018a].

Greater light input in exposed upper canopy leaves can sometimes increase $T_{leaf_{max}}$ beyond temperature optimum for photosynthesis ($T_{opt}$) and $T_{air}$ making upper canopy leaves susceptible to heat stress, more than scattered light in lower shaded leaves [@doughtyAreTropicalForests2008; @mauTemperateTropicalForest2018a]. Additionally, the upper canopy of temperate-deciduous species can have greater $T_{leaf}$ - $T_{air}$ than $T_{leaf}$ - $T_{air}$ of upper canopy tropical moist and wet forest species [@mauTemperateTropicalForest2018a].

@@

**Leaf thermoregulation is mediated by leaf traits. Leaf traits co-vary with their environment [@michaletzEnergeticCarbonEconomic2016], in order to optimize C uptake. Therefore, are critical to maintaining desirable $T_{leaf}$ and, in turn, shaping leaf metabolism (following section) across forest vertical strata (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`)[@michaletzPlantThermoregulationEnergetics2015; @zwienieckiHydraulicLimitationsImposed2004]**

#### *(clarify light vs height - 1 paragraph)*

Leaf morphological adaptation to the local microenvironement takes place during leaf expansion [@zwienieckiHydraulicLimitationsImposed2004]. Degrees of direct to diffused light and consequental water demands shape sun vs. shade leaves and their necessary vascular architecture [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a]. Since canopy leaves of taller trees are exposed to dramatically different biophysical conditions than leaves in the canopy interior or understory (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`) it is unsuprising, then, that leaf traits vary dramatically across vertical gradients.

While light-driven plasticity in leaf traits is well established, research on solely height-driven leaf trait variation is sparse due to the difficulty of isolating increased tree height from light. However, both theory and empirical evidence indicate that other factors--for example, height on a tree or exposure to evaporative demand (Fig. `r fig_NEON_vertical`)-- are important in shaping at least a subset of traits.

In particular, hydraulic constraints increase with height on a tree [REFS; *McDowell-isotopes book chapter*; @cobleLightAcclimationOptimizes2015; *Couvreur et al., 2018*] such that gas exchange is strongly constrained by at the tops of(~100 m. tall *Sequoia sempervirens*; *Ambrose et al. 2009, 2010*).
Isolating height (>35m) under constant light conditions, show that canopy stomatal conductance decreases along with increasing $VPD$  [@schaferEffectTreeHeight2000; @ambroseEffectsHeightTreetop2010]. Morphologically, leaf size [Kaare H. Jensen and Maciej A. Zwieniecki, 2013] and lobation [@kusiPlasticLeafMorphology2020] have shown to decrease, while leaf mass per area (at greater than 20m) to increase  [@cobleLightAcclimationOptimizes2015; @kenzoHeightrelatedChangesLeaf2015; @rijkersEffectTreeHeight2000a; @chinLeafAcclimationLight2017].

Disentangling the influences of hydraulic constraints, light, and other biophysical drivers that vary with height (Fig. `r fig_NEON_vertical`) on leaf form and function remains an important area for research, but is beyond the scope of the current review.
Our primary interest is in how leaf traits and function vary across the vertical gradient from the top of the canopy to the understory in forests.
For the purposes of interpreting the results presented here, we note that patterns across light and height gradients age generally similar, at least in direction (Table 1), but likely to be modified by covarying biophysical constraints.

@@

Within-individuals going from buffered [shaded] canopy positions to exposed, leaf traits vary on the basis of biophysical demands [@casasSunShadeLeaves2011] and therefore exhibit commonly observed patterns in leaf anatomical, structural and biochemical traits at an intra-canopy gradient, pointed by several studies [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020]. These observations contribute to [understanding] leaf temperature distribution across the gradient [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] Table `r table_leaf_traits` summarizes relevant traits for thermal sensitivity.

However, while a majority of relevant research have focused on exposure gradients near ground level, with by far the most common study type being comparisons of sun and shade leaves, given the vastly different light environments at the bottom of the canopy and that most leaves in the canopy are likely to be shade leaves, there is a strong research need to better characterize trait relationships and leaf responses through the whole vertical canopy [@keenanGlobalLeafTrait2016], such as by incorporating sun and shade in modelling leaf economic spectrum (LES) traits within-canopy gradient, partly because LES theory was developed using mainly sun exposed (upper canopy) leaves [@chenLeafEconomicsSpectrum2020].

@@

Going from shaded to exposed canopy position, morphological adaptation to dissapate heat increases. Sun leaves are often thicker, smaller, hairier, with greater dissection--lobes and serration--[@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @sackHowStrongIntracanopy2006], hydraulically efficient [Brodribb et al. 2010], with more vertical leaf angles than shade leaves [@niinemetsAdjustmentFoliageStructure1998]. All, collectively regulate $T_{leaf}$-photosynthetic balance. Under homogeneous light conditions in open-canopy, trait plasticity of sun leaves are based on $T_{air_{max}}$ and $VPD$ stress of canopy oriention [@mediavillaFoliarPlasticityRelated2019a]. Other hand, because shaded leaves have greater effective leaf width [@fauset_differencsunes_2018], boundary layer thickness, stomatal resistance and therefore longer thermal time constant, they exhibit lower convective cooling [@leighInfluenceLeafSize2017; Wright et al. 2019]. Which leads to cooler $T_{leaf}$ under buffered conditions, and substanially warmer $T_{leaf}$ under greater light or $T_{air}$ [@leighInfluenceLeafSize2017]

@@

Nitrogen distribution along the vertical gradient optimizes to prevent photoinhition in sun leaves during direct light and shade leaves during sunflecks under available soil nitrogen conditions [@kitaoCanopyNitrogenDistribution2018]. Therefore, during growing season, $N_{area}$, along with LMA, leaf soluble sugars, starch and non-structural carbons increase with exposure [@cobleLightDrivesVertical2014a; @cobleHowVerticalPatterns2016; Weerasinghe et al. 2014].

@@

Pools of xanthophyll pigments--violaxanthin, antheraxanthin and zeaxanthin [$VAZ$],an antioxidant scavenging function--plays a role in protecting against cellular damage by converting excess light into heat energy dissipation, enabling short-term acclimation to high temperature and light [@mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].Therefore, $VAZ$ increases proportional to vertical light gradients.

@@

##### *Leaf lifespan* [Elsa: phenology more broadly? - leaf lifespan, fruiting, flowering links to temperature and thermal gradients?]
*Leaf lifespan generally increases with LMA. LMA and leaflifespan is greater with available light, and largely depends on the cost of carbon investment in a leaf, which varies across species sun vs. shade leaves [**double check**] Leaf turnover tends to be greater in the sun (in the tropics), so of your standing stock of leaves (at least in evergreen species) sun leaves should be younger than shade leaves. This is bound to affect metabolism, and may affect other aspects of their performance.*

@@

**Similar to photosynthesis, respiration tends to be higher in sun leaves (Chen et al. 2020), but its temperature sensitivity appears to be similar between between sun and shade leaves.**

While respiration tends to increase along the vertical gradient, with greater respiration in canopy leaves [@bolstad_foliar_1999; Weerasinghe et al.2014], temperature sensitivity of respiration--$Q_{10}$--shows conflicting findings with no definite trend.

>>>>

**Light conditions, specifically changes in the proportion of incident light and photosynthetically active radiation (PAR), vary along the vertical gradient with leaf area index (LAI), canopy height, canopy structure, and across species and forest types [Fig. `r fig_NEON_vertical`; @koike_leaf_2001].** 
Generally, along a vertical gradient in forests globally, amount of light decreases from canopy tops to the forest floor. This trend is mediated by foliage LAI and density, where the gradient is more prounced in broad-leaf forests than needle-leaf forests [Fig. `r fig_NEON_vertical`; @aussenacInteractionsForestStands2000a; @vonarxSpatiotemporalEffectsForest2012a; Chazdon & Fetcher, 1984; Smith et al., 2019; Tymen et al., 2017]. Light reaching forest floors that is incident to canopy tops ranges from ~1-2% in tropical forests  [@robertsStomatalBoundaryLayerConductances1990a; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019], 4-5% in mixed-wood boreal and northern temperate forests[@bartemucciOverstoryInfluencesLight2006] to 17-34% in needle-leaf forests [@baldocchiSeasonalVariationEnergy1997]. Sunflecks, caused by small canopy gaps and wind-induced canopy movements also contribute to the shaded forest regions that receive 50-80% of their light for photosynthesis during these short exposures [@waySunflecksTreesForests2012]. Additionally, in heterogeneous canopies [with high gap fractions and large variation in tree height], the distance from the outer canopy to the inner canopy is a better proxy for light environment than height (Parker 1995).

### Wind speeds 
**Wind speeds are also higher at the top of the canopy (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`), owing to... [@juckerCanopyStructureTopography2018].**
Across all **#** forested NEON sites, wind speeds increased with height, with daily maximum wind speeds at the top of the averaging up to **#** m s^-1^ higher at the top of the vertical profile than at the bottom. Consistent with these results, higher wind speeds at greater heights have been observed in both closed-canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregor_tree_nodate] and open savannas [@curtis_intracanopy_2019].

@@

The strength of this buffering is greater under dense-canopy than open-canopy forests; with daytime cooler lower maximum  temperature, and nighttime warmer higher minimum in the understory, relatively more than open-forests, nearby clearings [*Defrenne et al. 2021*; @vonarxSpatiotemporalEffectsForest2012a, *Defrenne et al. 2021*; @davis_microclimatic_2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009] and neighboring tree crop agricultural plantations [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018]. On the other hand, open-forests maintain higher daytime maximum temperature in the understory closer to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a; @martin_boundary_1999; @ramboCanopyMicroclimateResponse2009] and  sometimes warmer nighttime understory temperatures than above canopy, due to long wave infrared radiation from the soil [@vonarxSpatiotemporalEffectsForest2012a].

Seasonally, buffering capacity increases with water availablity [@davisMicroclimaticBufferingForests2019] and is greater in wet than dry season [@rey-sanchez_spatial_2016; @vonarxSpatiotemporalEffectsForest2012a; @fauset_differences_2018, Tymen et al. 2017; @mcgregor_tree_nodate]. However, under colder conditions, dense-canopies maintain warmer higher minimum temperatures [$T_{min}$] in the understory relative to open-canopy forests and nearby clearings [@zellweger_seasonal_2019; *Defrenne et al. 2021*; @nakamuraForestsTheirCanopies2017; @mcgregor_tree_nodate] where greater radiative heat loss in exposed areas results in lower $T_{min}$.

Tree height also plays a role in buffering where trees upto 20m with higher LAI can maintain lower mean and max temperature and vapor pressure deficit ($VPD$) in the understory. However, beyond a certain tree height, temperature and $VPD$ offset can plateau [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018].

### Humidity and evaporative demand
**Humidity also varies across the forest vertical profile, and is generally higher in the understory (Fig. `r fig_NEON_vertical`).**[@mcgregor_tree_nodate]
Along with temperature, dense-canopy forests maintain higher daily maximum relative humidity ($RH$) below-canopy than open-forests and nearby open areas, which is greater in wetter conditions, in warmer than colder months [@vonarxSpatiotemporalEffectsForest2012a]. RH, in combination with temperature, determines $VPD$, an important metric expressing the driving force of water loss from a leaf; tends to be greater in upper-canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].

@@

With altitude, the strength of buffering decreases, moving to open-forests [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a; @rajsnerovaMorphologicalBiochemicalPhysiological2015; @liangForestTypeHeight2019], greater in wetter than drier regions [**REF**] and increases with proximity to water, for eg. in riparian forests. [@davisMicroclimaticBufferingForests2019; @macekMaximumAirTemperature2019]

@@

**The combination of biophysical variables reviewed above drive leaf temperature $T_{leaf}$ distribution along the vertical gradient, which in turn has a strong influence on leaf metabolism (see following section).**
Leaves adapt to their biophysical environment to maximize carbon assimilation, by maintaining optimum $T_{leaf}$ in a given circumstance [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020]. As a result, leaves can be cooler than air at high temperature and warmer than air at low temperature ('limited homeothermy') or can be drastically warmer than $T_{air}$ ('megathermy'). In some cases, $T_{leaf}$ can be equalled to $T_{air}$ ('poikilothermy') [@cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020].

$T_{leaf}$ itself is determined by the energy balance of a leaf, influenced by leaf traits, and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016] (Fig. `r fig_leaf_T`). The difference between $T_{leaf}$ and $T_{air}$ is determined as a function of energy input (through radiation or heat) versus heat lost (such as transpiration) to the environment [Fig. 2; @michaletzPlantThermoregulationEnergetics2015].

@@

Leaf thermoregulation is a tradeoff between water use strategy and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994]. Leaves can regulate their temperature by dissapating heat through stomatal conductance ($g_s$), even in drastic biophysical conditions [**REF**] but when leaf water demand exceeds water transport, driven by irradiance, $T_{air}$, VPD or drought, conservation of water ($gs$ limitation) occurs at the cost of increasing $T_{leaf}$.

@@

Greater wind speed (Fig. `r fig_NEON_vertical`) decreases $T_{leaf}$ - $T_{air}$ by increasing boundary layer conductance ($g_a$)  [@daudetWindSpeedLeaf1999a]. Therefore, under hot conditions, canopy-leaves exposed to higher wind speeds would be most effective at cooling with high $g_s$ [Drake et al. 2018]. At low wind conditions, small $g_a$ to heat transfer can increase $T_{leaf}$ - $T_{air}$ [@martin_boundary_1999; @leighInfluenceLeafSize2017] where boundary layer resistance increases with thickness of air layer at the leaf surface. This tends to be greater in buffered leaves [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990a; @stokesBoundaryLayerConductance2006]

**(paragraph on hydraulics as a driver of leaf T:e.g., water on leaves - evaporative cooling - and leaf water content)**
While stomatal conductance is inherently greater in exposed leaves than buffered shaded leaves, factors driving $g_s$ limitation--irradiance, VPD and $T_{air}$--are also greater with exposure. As a result, exposed leaves can experience steeper $g_s$ limitation than buffered leaves, driving an increase in $T_{leaf}$, further influenced by irradiance and $T_{air}$ (@leighInfluenceLeafSize2017; @fauset_differences_2018). On the other hand, due to inherently lower $g_s$ in shade leaves, they can remain cooler under buffered conditions and substantially warmer under increased $T_{air}$ or irradiance.

@@

### In-situ observations
**Vertical profiles in microclimate (Fig. `r fig_NEON_vertical`), leaf traits (see following section), and *evapotranspiration* (*where is this discussed? and do we say anything about evaporation?*) together shape in-situ leaf temperatures according to the principles outlined above (Fig. `r fig_leaf_T`).** 
Canopy buffering capacity largely influences vertical $T_{leaf}$ dynamics. Particularly, in open-forests with lower buffering capacity, $T_{leaf}$ is greater in lower-canopy than upper canopy, reflecting microclimate conditions of vertical air mixing, light permeating through canopy layers, still air in the lower canopy and heating from the ground, [@mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987;@rey-sanchez_spatial_2016]. Similarly, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019]. Greater $RH$ in inner canopy can also increase $T_{leaf}$  relatively more than upper-canopy $T_{leaf}$ [@dietzVerticalPatternsDuration2007]. Under other conditions $T_{leaf}$ tracks $T_{air}$ and therefore differs little between understory and canopy (Fig. `r fig_NEON_vertical`).

**However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.** Closed-canopy forests, such as tropical and semi-deciduous tropical, and temperate forests, experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies due to greater heat and water stress from $gs$ limitation, than buffered lower-canopies [@doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016]. Greater light input in exposed upper-canopy leaves can sometimes increase $T_{leaf_{max}}$ beyond temperature optimum for photosynthesis ($T_{opt}$) and $T_{air}$ making upper-canopies susceptible to heat stress [@doughtyAreTropicalForests2008; @mauTemperateTropicalForest2018a]. 
Comparing forests, upper-canopies of temperate-deciduous species can have greater $T_{leaf}$ - $T_{air}$ than the upper-canopies of tropical moist and wet forest species [@mauTemperateTropicalForest2018a].

@@

**Leaf thermoregulation largely depends leaf traits, that are in-turn adapted to their environment [@michaletzEnergeticCarbonEconomic2016]. Therefore, are critical in maintaining desirable $T_{leaf}$ and in shaping leaf metabolism (following section) across forest vertical strata (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`)[@michaletzPlantThermoregulationEnergetics2015; @zwienieckiHydraulicLimitationsImposed2004]**

Leaf morphological adaptation to the local microenvironement takes place during leaf expansion [@zwienieckiHydraulicLimitationsImposed2004]. Degrees of direct to diffused light and resultant water demands shape sun vs. shade leaves and their necessary vascular architecture [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a]. Therefore, taller tree canopy-leaves dramatically vary from canopy-interior and understory due to the differences in biophysical conditions (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`). 
While leaf traits vary primarily on the axis of light, at greater tree-heights, both theory and empirical evidence indicate that other factors-for example, tree-height or exposure to evaporative demand (Fig. `r fig_NEON_vertical`)- are also important in shaping at least a subset of traits.

In particular, hydraulic constraints increase with height on a tree [REFS; *McDowell-isotopes book chapter*; @cobleLightAcclimationOptimizes2015; *Couvreur et al., 2018*] such that gas exchange is strongly constrained by at the tops of(~100 m. tall *Sequoia sempervirens*; *Ambrose et al. 2009, 2010*).
Isolating height (>35m) under constant light conditions show that, canopy $gs$ decreases along with increasing $VPD$  [@schaferEffectTreeHeight2000; @ambroseEffectsHeightTreetop2010]. Morphologically, leaf size [Kaare H. Jensen and Maciej A. Zwieniecki, 2013] and lobation [@kusiPlasticLeafMorphology2020] have shown to decrease, while leaf mass per area (at greater than 20m) to increase  [@cobleLightAcclimationOptimizes2015; @kenzoHeightrelatedChangesLeaf2015; @rijkersEffectTreeHeight2000a; @chinLeafAcclimationLight2017].

Disentangling the influences of hydraulic constraints, light and other biophysical drivers that vary with height (Fig. `r fig_NEON_vertical`) on leaf traits remains an important area for research, but is beyond the scope of the current review. Our primary interest is in how leaf traits and function vary across the vertical gradient from the top of the canopy to the understory in forests. For the purposes of interpreting the results presented here, we note that patterns across light and height gradients are generally similar, at least in direction (Table 1), but likely to be modified by covarying biophysical constraints.

@@

Within-canopy gradient of individuals, leaf traits vary--going from buffered [shaded] to exposed canopy positions [@casasSunShadeLeaves2011]-- and therefore exhibit patterns in leaf anatomical, structural and biochemical traits [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020]. These trait-patterns contribute to [understanding] $T_{leaf}$ dynamics across the gradient [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] where Table `r table_leaf_traits` summarizes relevant traits for thermal sensitivity. The characterization of trait relationships and leaf responses through the whole vertical canopy is becoming a strong research need because LES theory was developed using mainly sun exposed (upper canopy) leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016].

@@

Going from shaded to exposed canopy position, morphological adaptation to dissapate heat increases. Sun leaves are often thicker, smaller, hairier, with greater dissection--lobes and serration--[@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @sackHowStrongIntracanopy2006], hydraulically efficient [Brodribb et al. 2010], with more vertical leaf angles than shade leaves [@niinemetsAdjustmentFoliageStructure1998]. All, collectively regulate $T_{leaf}$-photosynthetic balance. On the other hand, shaded leaves invest more in $SLA$ for light capture than in heat dissapation, therefore have lower convective cooling [@fauset_differencsunes_2018; @leighInfluenceLeafSize2017; Wright et al. 2019]. Additionally, under homogeneous light conditions, sun-leaves in different canopy orientations are shaped on the basis of $T_{air_{max}}$ and $VPD$ stress [@mediavillaFoliarPlasticityRelated2019a].

@@

Nitrogen along the vertical gradient, is constantly re-mobilized from senescensed to growing leaves [@niinemetsPhotosynthesisResourceDistribution2007a] where $N_{area}$, with LMA, leaf soluble sugars, starch and non-structural carbons are greater [@cobleLightDrivesVertical2014a; @cobleHowVerticalPatterns2016; Weerasinghe et al. 2014]. Additionally, $N_{area}$ optimizes to prevent photoinhition in sun leaves during direct light and shade leaves during sunflecks, under available soil nitrogen [@kitaoCanopyNitrogenDistribution2018].

@@

Pools of xanthophyll pigments--violaxanthin, antheraxanthin and zeaxanthin [$VAZ$], an antioxidant scavenging function--plays a role in converting excess light into heat energy dissipation, enabling short-term acclimation to high temperature and light [@mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998], increases proportional light gradients.

##### Isoprene production
Isoprene production *capability* increases vertically with light and therefore facilitates high heat tolerance in emitting species' canopy positons [Sharky]

Across tree-size ontogeny of the same species, while fundamental adaptations to light and water are observed in canopy and understory trees similar to within-species gradient [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012], increase in LMA, photosynthetic capcity per area, and stomatal conductance with height are attributed to ontogenetical differences, under constant light conditions [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012]. Across  multi-species canopy and understory, trait adaptation to spatio-temporal light and biophysical gradient varies largely with inherent specie-specific trait differences such as leaf shape, size, $gs$ levels and therefore influences individual $T_{leafs}$. As a result, some species might be more susceptible to heat stress at a vertical gradient than others [@fauset_differences_2018].

@@

##### *Leaf lifespan* 
*Leaf turnover tends to be greater in the sun (in the tropics), so of your standing stock of leaves (at least in evergreen species) sun leaves should be younger than shade leaves. This is bound to affect metabolism, and may affect other aspects of their performance.*

@@

**Similar to photosynthesis, respiration tends to be higher in canopy sun-leaves (Chen et al. 2020), but temperature sensitivity of respiration--$Q_{10}$--shows no definite trend along the gradient [@bolstad_foliar_1999; Weerasinghe et al.2014].**
